--- 
title: "Practical R Applications in Water Resources"
author: "Michael Schramm"
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
documentclass: book
bibliography: [book.bib, packages.bib]
biblio-style: apalike
link-citations: yes
description: "This is a DRAFT!!!"
---

# Prerequisites {-}

This is not intended to be a primer or introduction to programming or R. Instead, this is a cookbook of sorts to help guide users to applying R in the context of our water resources work. External resources for those interested in the basics are provided below ([Familiarity with R and RStudio](#basics)).
A few things required before getting started...

## Install R {-}

Download and install the newest version of R for your platform at: [https://cloud.r-project.org/](https://cloud.r-project.org/). For *Windows*:

- click on "Download R for Windows"
- click on "base"
- click the download link at the top of the page
- open the downloaded `R-4.x.x.exe` file and follow the prompts

For *macOS*:

- click on "Download R for (Mac) OS X"
- Under "Latest release:" click on `R-4.x.x.pkg`

For *Linux*:

- click "Download R for Linux" and choose the appropriate distribution


## Install RStudio {-}

Install the latest version of Rstudio Desktop for your platform at:
[https://rstudio.com/products/rstudio/download/#download](https://rstudio.com/products/rstudio/download/#download)


## Familiarity with R and RStudio {-#basics}

This text assume some basic familiarity with R and RStudio. At minimum you should know how to start a project in RStudio, have a basic understanding of R packages and functions...


If you are completely new to using R I recommend a few resources below:

[https://tinystats.github.io/teacups-giraffes-and-statistics/](https://tinystats.github.io/teacups-giraffes-and-statistics/) - Great if you are completely new to programming, also provides some foundational stats training.

[https://rstudio.cloud/learn/primers](https://rstudio.cloud/learn/primers) - This provides self-paced, interactive, fundamental R programming exercises.

[Adventures in R](https://www.adventures-in-r.com/post/01-working-in-r/) - Another self-paced set of videos and excercies. This has a good introduction to the RStudio IDE.


## Suggested Packages {-}

This text will use the `tidyverse` set of packages throughout.

```{r eval=FALSE}
install.pacakges("tidyverse")
```


Sections on spatial analysis will utilize `sf` and `raster`.
```{r eval=FALSE}
install.pacakges(c("sf", "raster"))
```

## Contributions and Editing {-}

This text is rendered using `bookdown` and GitHub Actions [@bookdown2016]. This means you can submit issues, add code, or copy and render the book on your own system. The project is located on [GitHub](https://github.com/mps9506/Tutorials). If you clone the project into your RStudio session, the book can be rendered as local website using:

```{r eval=FALSE}
bookdown::render_book("index.Rmd")
```

Navigate to the `\_book` subdirectory and open `index.html` to see the book.

Furthermore, a pdf copy of the book can be generated using:
```{r eval=FALSE}
bookdown::render_book("index.Rmd", "bookdown::pdf_book")
```

Navigate to the `\_book` subdirectory and open `book.pdf`.
